An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-03-12 EconGrowth-R-Functions.Rnw
Now correctly calculating beta for cde models. Also, developing graphs of SSE, lambda, alpha, and beta vs. gamma to show that the minimum SSE occurs at the point where beta --> 0.0.

###############################
# Does a grid search over values of gamma for a CD production function
##
cdeGridData <- function(countryAbbrev, energyType, gammaGrid){
  gamma <- gammaGrid
  models <- lapply(gamma, cobbDouglasModel, energyType=energyType, countryAbbrev=countryAbbrev)
  return(models)
}

#############################
# This function creates a series of graphs for Cobb-Douglas with energy functions
# showing various parameters as a function of gamma.
##
cdeGridGraphs <- function(countryAbbrev, energyType){
  gammaGrid <- seq(0.0, 0.95, by=0.05)
  #print(paste(countryAbbrev, energyType))
  models <- cdeGridData(countryAbbrev, energyType, gammaGrid)
  SSE <- lapply(models, function(model) summary(model)$sigma)
  lambda <- lapply(models, function(model) coef(model)["lambda"])
  alpha <- lapply(models, function(model) coef(model)["alpha"])
  data <- cbind(lambda, alpha, gammaGrid, SSE)
  plot(gammaGrid, SSE)
  plot(gammaGrid, lambda)
  plot(gammaGrid, alpha)
}

@@

    betaCICD <- with(beta.est, Estimate + c(-1.0, 1.0) * tvalCD * SE); betaCICD # Gives the confidence interval on beta.  

>>>>

###############################
# Does a grid search over values of gamma for a CD with energy production function.
# Results are returned as a data.frame with columns of gamma, SSE, lambda, alpha, and beta
##
cdeGridData <- function(countryAbbrev, energyType, gammaGrid){
  gamma <- gammaGrid
  models <- lapply(gamma, cobbDouglasModel, energyType=energyType, countryAbbrev=countryAbbrev)
  SSE <- unlist(lapply(models, function(model){summary(model)$sigma}))
  lambda <- unlist(lapply(models, function(model){coef(model)["lambda"]}))
  alpha <- unlist(lapply(models, function(model){coef(model)["alpha"]}))
  data <- as.data.frame(cbind(gammaGrid, lambda, alpha, SSE))
  row.names(data) <- NULL #eliminates row names, leaving only numbered rows
  colnames(data) <- c("gamma", "lambda", "alpha", "SSE")
  data <- transform(data, beta = 1.0 - alpha - gamma)
  return(data)
}

#############################
# This function creates a series of graphs for Cobb-Douglas with energy functions
# showing various parameters as a function of gamma.
##
cdeGridGraphs <- function(countryAbbrev, energyType){
  gammaGrid <- seq(0.0, 0.95, by=0.05)
  data <- cdeGridData(countryAbbrev, energyType, gammaGrid)
  with(data, plot(gamma, SSE))
  with(data, plot(gamma, lambda))
  with(data, plot(gamma, alpha))
  with(data, plot(gamma, beta))
  with(data, plot(beta, SSE))
}

@@

### 2013-03-12 EconGrowth.Rnw
Now correctly calculating beta for cde models. Also, developing graphs of SSE, lambda, alpha, and beta vs. gamma to show that the minimum SSE occurs at the point where beta --> 0.0.

<<CDeGammaGraphsZA>>=
  cdeGridGraphs("ZA", "Q")
@

@@

<<LINEX_GDP_Lattice_Graph, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.pos="H", fig.cap="LINEX results.">>=
createLinexLatticeGraph()
@

>>>>

<<CDeGammaGraphsZA, eval=TRUE>>=
  cdeGridGraphs("ZA", "Q")
@

@@

<<LINEX_GDP_Lattice_Graph, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.pos="H", fig.cap="LINEX results.">>=
createLINEXLatticeGraph()
@

### 2013-03-12 EconGrowthSeminar.Rnw
Added Bashmakov reference.
Now correctly calculating beta for cde models. Also, developing graphs of SSE, lambda, alpha, and beta vs. gamma to show that the minimum SSE occurs at the point where beta --> 0.0.

\usepackage{times}
\usepackage[T1]{fontenc}
% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.
\usepackage{url} % To format URLs in the references section.

@@

\begin{frame}{Energy and GDP are Linked} %----------------------------
  \tiny
  \framesubtitle{\cite{Cleveland:1997p5055}}
  \normalsize
  \begin{columns}[c]
    \column{2.0in}
      \includegraphics[width=2.0in]{figure_other/Figure_Cleveland_1984.pdf}
    \column{1.5in}
      \includegraphics[width=1.5in]{figure_other/Figure_Cleveland_1984_Science_Cover.pdf}
  \end{columns}
\end{frame}

\begin{frame}{World Oil Supply is Tightening} %----------------------------
  \framesubtitle{\cite{Tverberg:2013}}
  \includegraphics[width=4.0in]{figure_other/Figure_Tverberg_2013.png}
\end{frame}

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item
      Therefore, energy is considered to be economically unimportant.
    \item
      Compare cost shares of capital stock, labor, and energy.
  \end{itemize}
\end{frame}

\begin{frame}{Energy Cost Share and Economic Growth} %----------------------------
  \begin{center}
    \includegraphics[height=6.3cm]{figure_other/Figure_Bashmakov_2007.pdf}
  \end{center}
%  \citet{Bashmakov:2007ek}
\end{frame}

\begin{frame}{Observations and Questions} %----------------------------
  \begin{itemize}
    \item
      Historical economic development paths are energy intensive. 
    \item
      What happens when energy supplies are constrained and prices rise? 
    \item
      Will developing countries be able to expand their economies?
    \item
      How account for energy in economic models?
  \end{itemize}
\end{frame}

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Figure_Growth_Models.pdf}
  \end{center}
  \begin{itemize}
  \item
    Goal: predict economic output over time
  \item
    Economic output is a function of ``factors of production''
  \item
    Output defined as GDP ($Y$)
  \item
    All inputs and outputs normalized to initial year (0)
  \end{itemize}
\end{frame}

@@

\begin{frame}{Capital Stock ($K$) and Labor ($L$)} %----------------------------
  \begin{itemize}
  \item
    Capital Stock ($K$) <Definition here>
  \item
    Labor ($L$) <Definition here>
  \end{itemize}
\end{frame}

@@

\begin{frame}{Single Factor} %----------------------------

  \begin{equation*} \label{eq:Single_Factor_Generic}
    y = \mathrm{e}^{\lambda (t-t_0)}f^{m}
  \end{equation*}

  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $y$   & $Y$/$Y_0$ \\
      $t$   & year\\
      $t_0$ & initial year\\
      $f$   & $F$/$F_0$, a factor of production\\
      \hline
      $\lambda$   &   technological progress\\
      $m$   &         ``strength'' of factor $f$\\
      \hline
    \end{tabular}
  \end{center}

@@

\begin{frame}{Cobb-Douglas} %----------------------------

  \begin{equation*} \label{eq:CD_With_Energy}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}
  \end{equation*}

  \begin{equation*} \label{eq:CD_Const_Returns_To_Scale}
    \alpha + \beta + \gamma = 1
  \end{equation*}

  \begin{equation*} \label{eq:CD_Constraints}
    0 \leq \alpha, \beta, \gamma \leq 1
  \end{equation*}

  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\lambda$   &   technological progress\\
      $\alpha$   &    ``strength'' of capital\\
      $\beta$   &    ``strength'' of labor\\
      $\gamma$   &    ``strength'' of energy\\
      \hline
    \end{tabular}
  \end{center}

@@

\begin{frame}{Constant Elasticity of Substitution (CES)} %----------------------------

  \begin{equation*} \label{eq:micEconCES_model}
    \begin{split}
    y = & \gamma \mathrm{e}^{\lambda(t-t_0)} \\
        & \cdot \left\{\delta \left[\delta_1 k^{-\rho_1} + (1-\delta_1)l^{-\rho_1} \right]^{\rho/\rho_1} 
                        + (1-\delta) e^{-\rho} \right\}^{-1/\rho}
    \end{split}
  \end{equation*}

  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\gamma$   &   fitting coefficient\\
      $\lambda$  &   technological progress\\
      $\delta$   &   split between ($kl$) and ($e$)\\
      $\rho$     &   ``strength'' of ($kl$) vs. ($e$)\\
      $\delta_1$   &   split between ($k$) and ($l$)\\
      $\rho_1$     &   ``strength'' of ($k$) vs. ($l$)\\
      \hline
    \end{tabular}
  \end{center}

@@

\begin{frame}{LINEX} %----------------------------

  \begin{equation*} \label{eq:LINEX_model}
    y = e \: \mathrm{e}^{\left[a_0 \left(2 - \frac{l + e}{k} \right) + a_0 c_t \left(\frac{l}{e} - 1 \right)\right]}
  \end{equation*}

  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $a_0$   &   fitting parameter\\
      $c_t$   &   fitting parameter\\
      \hline
    \end{tabular}
  \end{center}

@@

@@

% See hints at http://judsonsnotes.com/notes/index.php?option=com_content&view=article&id=641:bibliography-in-beamer&catid=60:latex&Itemid=84
% \begin{frame}{Bibliography} %----------------------------
% \bibliographystyle{model2-names}
% \bibliography{EconGrowth.bib}
% \end{frame}

%-------------------------------------------------------------------

\begin{frame}
\frametitle{References}
%\bibliographystyle{amsalpha}
\bibliographystyle{model2-names}
\renewcommand{\thepage}{\arabic{page}}
\tiny
\bibliography{EconGrowth.bib}
\normalsize
\end{frame}

>>>>

\usepackage{times}
\usepackage[T1]{fontenc}
% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.
\usepackage{url} % To format URLs in the references section.
\usepackage{natbib} % To format the bibliobraphy nicely

@@

\begin{frame}{Energy and GDP are Linked} %----------------------------
  \tiny
  \framesubtitle{\citep{Cleveland:1997p5055}}
  \normalsize
  \begin{columns}[c]
    \column{2.0in}
      \includegraphics[width=2.0in]{figure_other/Figure_Cleveland_1984.pdf}
    \column{1.5in}
      \includegraphics[width=1.5in]{figure_other/Figure_Cleveland_1984_Science_Cover.pdf}
  \end{columns}
\end{frame}

\begin{frame}{World Oil Supply is Tightening} %----------------------------
  \framesubtitle{\citep{Tverberg:2013}}
  \includegraphics[width=4.0in]{figure_other/Figure_Tverberg_2013.png}
\end{frame}

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item{Energy is considered to be economically unimportant.}
    \item{Compare cost shares of capital stock, labor, and energy.}
  \end{itemize}
\end{frame}

\begin{frame}{Energy Cost Share and Economic Growth} %----------------------------
  \framesubtitle{\citep{Bashmakov:2007ek}}
  \begin{center}
    \includegraphics[height=6.3cm]{figure_other/Figure_Bashmakov_2007.pdf}
  \end{center}
%  \citet{Bashmakov:2007ek}
\end{frame}

\begin{frame}{Observations and Questions} %----------------------------
  \begin{itemize}
    \item{Historical economic development paths (US, Western Europe) are energy intensive.}
    \item{What happens when energy supplies are constrained and prices rise?}
    \item{Will developing countries be able to expand their economies?}
    \item{How account for energy in economic models?}
  \end{itemize}
\end{frame}

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Figure_Growth_Models.pdf}
  \end{center}
  \begin{itemize}
  \item{Goal: predict economic output over time}
  \item{Economic output is a function of ``factors of production''}
  \item{Output defined as GDP ($Y$)}
  \item{All inputs and outputs normalized to initial year (0)}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Capital Stock ($K$) and Labor ($L$)} %----------------------------
  \begin{itemize}
  \item{Capital Stock ($K$) <Definition here>}
  \item{Labor ($L$) <Definition here>}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Single Factor} %----------------------------
  \begin{equation*} \label{eq:Single_Factor_Generic}
    y = \mathrm{e}^{\lambda (t-t_0)}f^{m}
  \end{equation*}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $y$   & $Y$/$Y_0$ \\
      $t$   & year\\
      $t_0$ & initial year\\
      $f$   & $F$/$F_0$, a factor of production\\
      \hline
      $\lambda$   &   technological progress\\
      $m$   &         ``strength'' of factor $f$\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

\begin{frame}{Cobb-Douglas} %----------------------------
  \begin{equation*} \label{eq:CD_With_Energy}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}
  \end{equation*}
  \begin{equation*} \label{eq:CD_Const_Returns_To_Scale}
    \alpha + \beta + \gamma = 1
  \end{equation*}
  \begin{equation*} \label{eq:CD_Constraints}
    0 \leq \alpha, \beta, \gamma \leq 1
  \end{equation*}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\lambda$   &   technological progress\\
      $\alpha$   &    ``strength'' of capital\\
      $\beta$   &    ``strength'' of labor\\
      $\gamma$   &    ``strength'' of energy\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

\begin{frame}{Constant Elasticity of Substitution (CES)} %----------------------------
  \begin{equation*} \label{eq:micEconCES_model}
    \begin{split}
    y = & \gamma \mathrm{e}^{\lambda(t-t_0)} \\
        & \cdot \left\{\delta \left[\delta_1 k^{-\rho_1} + (1-\delta_1)l^{-\rho_1} \right]^{\rho/\rho_1} 
                        + (1-\delta) e^{-\rho} \right\}^{-1/\rho}
    \end{split}
  \end{equation*}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\gamma$   &   fitting coefficient\\
      $\lambda$  &   technological progress\\
      $\delta$   &   split between ($kl$) and ($e$)\\
      $\rho$     &   ``strength'' of ($kl$) vs. ($e$)\\
      $\delta_1$   &   split between ($k$) and ($l$)\\
      $\rho_1$     &   ``strength'' of ($k$) vs. ($l$)\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

\begin{frame}{LINEX} %----------------------------
  \begin{equation*} \label{eq:LINEX_model}
    y = e \: \mathrm{e}^{\left[a_0 \left(2 - \frac{l + e}{k} \right) + a_0 c_t \left(\frac{l}{e} - 1 \right)\right]}
  \end{equation*}
  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $a_0$   &   fitting parameter\\
      $c_t$   &   fitting parameter\\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

@@

\begin{frame}{Next Steps (Summer 2013)} %----------------------------
  \begin{itemize}
    \item{Estimate statistical properties for \emph{all} fitting parameters}
    \item{Complete paper}
  \end{itemize}  
\end{frame}

@@

\begin{frame}{References} %----------------------------
  \bibliographystyle{plainnat}
  \renewcommand{\thepage}{\arabic{page}}
  \tiny
    \bibliography{EconGrowth.bib}
  \normalsize
\end{frame}
